The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-09 paper/iphas-dr2.tex
First set of fixes for comments from CK

Over the years it has become apparent that the legacy of IPHAS 
will also reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed however, 
it is vital that the imaging data are transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we briefly recall some key points.
IPHAS is a 1800~deg$^2$. imaging survey of the northern Galactic Plane, 
providing images and photometry in Sloan $r, i$ 
along with narrowband H$\alpha$. 
It is carried out using the WFC 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive CCD photometry
of point sources in the Galactic Plane at visible wavelengths, 
and does so down to a limiting magnitude of $\sim$20th.
The IPHAS footprint spans a box 
of roughly 180 by 10 degrees, 
taking in the entire northern Galactic Plane 
at latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only the offset position is missing).}
        \label{fig:footprint}
\end{figure*}

@@

All raw data obtained with the INT were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}
and much of the source code is available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the traditional time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that was taken as a result
of this strategy has subsequently been flagged and removed
using a series of seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations fared significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations and explains the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depth attained in $i$ and \ha\ 
is less sensitive to moonlight
and their magnitude limit distributions
are hence more narrow
(middle and bottom panels in Fig.~\ref{fig:depth}).

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\% of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at the $\sim$same time --  
essentially always within an elapsed time of 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr}
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(typically 0.09 for $r$ and 0.05 for $i$ in La Palma),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used.
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs, 
and as such is an ancillary correction 
to account for sporadic gain variations. 
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

@@

Unlike the broadbands, 
standard star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for the narrowband.
We can specify here however 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter throughput curve 
corrected for the atmosphere and detector quantum efficiency
\citep[following the method explained by][]{Drew2005}.
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}

@@

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
Firstly, the survey was carried out 
in competitive observing time
on a non-dedicated telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive
-- in part because standard fields 
are very scarce within the Galactic Plane.
Secondly, the aim of IPHAS is to obtain magnitudes at a single epoch
and hence stars are not normally observed at more than one epoch,
unless they happen to fall within a narrow overlap region 
between two neighbouring fields.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

@@

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2
   \label{eqn:chi2}
\end{equation}
where $i$ denotes the exposure of interest, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the uncertainty in $\Delta_{ij}$
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$ which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}
This prescription is essentially identical to \citet{Glazebrook1994}.
%,although we changed the formulation to ensure that $A_{ij}$ is positive definite.

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude.
Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error
and we do not recommend its use.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band and hence the $(r-i)$ and $(r-\ha)$ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
Such very bright stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
We will explain below that no such detections 
have been included in the catalogue.
An alternative detection was available in essentially all these situations
as a result of the IPHAS field pair strategy,
and hence it was not necessary to include these internal warning flags
in the source catalogue.

@@

The resulting band-merged catalogues were found
to be reliable for the vast majority of fields.
We do warn that source blending and confusion is unavoidable
for faint objects in the most crowded regions of the Galactic Plane;
in \S\ref{sec:discussion} we will show
that 19\% of the sources in our catalogue
are flagged as blended objects (column \emph{deblend})
and their band-merged data should be treated with care
because they may have fallen victim to source confusion
during the band-merging step.

@@

Each of the sight-lines reveal a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre 
at $l=180^\circ$ (Fig.~\ref{fig:l180})
we find a population dominated by lowly-reddened main sequence stars.
This is consistent with the estimated total sight-line extinction 
of $E(B-V)=0.49$ given by \cite{Schlegel1998},
following the correction of \cite{Schlafly2011}.
Looking in more detail we can see
that the stellar locus is narrower for M-type dwarfs
than for earlier types;
we do not observe M dwarfs experiencing 
the strongest reddening possible for this sight-line.
This implies that extinction is still increasing 
at distances beyond $\sim$1-2~kpc,
where M dwarfs are too faint to be contained in the IPHAS catalogue.
It is also clear that there are no unreddened stars
earlier than $\sim$K0 visible,
such stars would be saturated if within a few hundred parsecs.
This therefore suggests 
that there is a measurable increase in extinction locally.
We also note a relative absence of late type giants which,
due to the relative brevity 
of the corresponding phase of stellar evolution,
would only account for a small proportion of a volume limited sample.

In contrast, 
closer towards the galactic centre
at $l=45^{\circ}$ (Fig.~\ref{fig:l45})
we find a wealth of reddened objects.
In the colour-magnitude diagram it is clear
that the stars are split into two distinct groups,
with one significantly redder than the other.
The bluer group is composed of main sequence stars,
with the slope of this group in the colour-magnitude diagram
attributable to the significantly increasing extinction.
Meanwhile the redder group is principally composed of red giant stars.
As these stars are intrinsically brighter,
at some given apparent magnitude they will be substantially further away
than their main sequence counterparts.
Given that, along this sight-line,
extinction continues to increase with distance,
the red giants we observe will be subject to more reddening
than a main sequence star exhibiting the same apparent magnitude.
Therefore, the clear split between the two groups arises
due to the combination of increasing extinction
and the distinct absolute magnitude ranges
of main sequence and giant branch stars.

>>>>

Over the years it has become apparent that the legacy of IPHAS 
will also reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we briefly recall some key points.
IPHAS is a 1800~deg$^2$. imaging survey of the northern Galactic Plane, 
providing images and photometry in Sloan $r, i$ 
along with narrowband H$\alpha$. 
It is carried out using the WFC 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive CCD photometry
of point sources in the Galactic Plane at visible wavelengths, 
and does so down to a limiting magnitude of $\sim$20.
The IPHAS footprint spans a box 
of roughly 180 by 10 degrees, 
taking in the entire northern Galactic Plane 
at latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only the offset position is missing).
        The dark vertical strip near $l \sim$125\deg
        is an uninteresting consequence of the tiling pattern,
        which overlaps in the middle of the survey area by design.}
        \label{fig:footprint}
\end{figure*}

@@

All raw data obtained with the INT were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and much of the source code is available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the traditional time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that were taken as a result
of this strategy has subsequently been flagged and removed
using a series of seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations fared significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations and explains the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight,
and their magnitude limit distributions
are hence more narrow
(middle and bottom panels in Fig.~\ref{fig:depth}).

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec,
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\% of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at roughly the same time --  
essentially always within an elapsed time of 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(typically 0.09 for $r$ and 0.05 for $i$ in La Palma),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used.
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs, 
and as such is an ancillary correction 
to account for sporadic gain variations. 
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

@@

Unlike the broadbands, 
standard star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for the narrowband.
We can specify here, however, 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter throughput curve 
corrected for the atmosphere and detector quantum efficiency
\citep[following the method explained by][]{Drew2005}.
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}

@@

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
Firstly, the survey was carried out 
in competitive observing time
on a non-dedicated telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive
-- in part because standard fields 
are very scarce within the Galactic Plane.
Secondly, the aim of IPHAS is to obtain magnitudes at a single epoch,
and hence stars are not normally observed at more than one epoch,
unless they happen to fall within a narrow overlap region 
between two neighbouring fields.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS,
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

@@

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2
   \label{eqn:chi2}
\end{equation}
where $i$ denotes the exposure of interest, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the inverse square of the uncertainty in $\Delta_{ij}$
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$, which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j,
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}
This prescription is essentially identical to \citet{Glazebrook1994}.
%,although we changed the formulation to ensure that $A_{ij}$ is positive definite.

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude.
Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error,
and we do not recommend its use.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band, and hence the $(r-i)$ and $(r-\ha)$ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
Such very bright stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
We will explain below that no such detections 
have been included in the catalogue.
An alternative detection was available in essentially all these situations
as a result of the IPHAS field pair strategy,
and hence it was not necessary to include these internal warning flags
in the source catalogue.

@@

The resulting band-merged catalogues were found
to be reliable for the vast majority of fields.
We do warn that source blending and confusion is unavoidable
for faint objects in the most crowded regions of the Galactic Plane;
in \S\ref{sec:discussion} we will show
that 19\% of the sources in our catalogue
are flagged as blended objects (column \emph{deblend}),
and their band-merged data should be treated with care
because they may have fallen victim to source confusion
during the band-merging step.

@@

Each of the sight-lines reveal a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre 
at $l=180^\circ$ (Fig.~\ref{fig:l180})
we find a population dominated by lowly-reddened main sequence stars.
This is consistent with the estimated total sight-line extinction 
of $E(B-V)=0.49$ given by \cite{Schlegel1998},
following the correction of \cite{Schlafly2011}.
Looking in more detail we can see
that the stellar locus is narrower for M-type dwarfs
than for earlier types;
we do not observe M dwarfs experiencing 
the strongest reddening possible for this sight-line.
This implies that extinction is still increasing 
at distances beyond $\sim$1-2~kpc,
where M dwarfs are too faint to be contained in the IPHAS catalogue.
It is also clear that there are no unreddened stars
earlier than $\sim$K0 visible;
such stars would be saturated if within a few hundred parsecs.
This therefore suggests 
that there is a measurable increase in extinction locally.
We also note a relative absence of late type giants which,
due to the relative brevity 
of the corresponding phase of stellar evolution,
would only account for a small proportion of a volume limited sample.

In contrast, 
closer towards the galactic centre
at $l=45^{\circ}$ (Fig.~\ref{fig:l45}),
we find a wealth of reddened objects.
In the colour-magnitude diagram, it is clear
that the stars are split into two distinct groups,
with one significantly redder than the other.
The bluer group is composed of main sequence stars,
with the slope of this group in the colour-magnitude diagram
attributable to the significantly increasing extinction.
Meanwhile the redder group is principally composed of red giant stars.
As these stars are intrinsically brighter,
at some given apparent magnitude they will be substantially further away
than their main sequence counterparts.
Given that, along this sight-line,
extinction continues to increase with distance,
the red giants we observe will be subject to more reddening
than a main sequence star exhibiting the same apparent magnitude.
Therefore, the clear split between the two groups arises
due to the combination of increasing extinction
and the distinct absolute magnitude ranges
of main sequence and giant branch stars.
